---
title: "ICD10"
output: html_document
---

# Startup

The goal of this notebook is to create a one-stop shop for medical codings in the UK Biobank. This was partially done by the UK Biobank via the first occurrence variable.

```{r startup}

library(data.table)
library(stringr)
library(tidyr)
library(lubridate)

# reticulate is not installed by default and is required by the python interpreter in RStudio
if (!require(reticulate)) install.packages('reticulate')

```

# Download Requisite Files

See python notebooks in the folders below for how data was downloaded from the UKBB phenotype database. Fields included are:

* First Occurrence Data (fields 130000 - 132605, not inclusive)
* Hospital Episode Statistics
    * ICD-10 (41270)
    * ICD-9 (41271)
* Cancer Registry
    * ICD-10 (40006)
    * ICD-9 (40013)
* Death Registry
    * Primary ICD10 (40001)
    * Secondary ICD10 (40002)

**Note**: I have not retrieved self-reported conditions as they are presumably better mapped to specific ICD-10 code by the
team that built the first occurrence field(s).

For ICD9 -> ICD10 mappings, I downloaded the file provided by the US CMS from this webpage and uploaded to DNANexus:

https://www.cms.gov/Medicare/Coding/ICD10/2018-ICD-10-CM-and-GEMs

```{bash dx downloads}

# Download summary tables
dx download phenotypes/summary_diagnoses/40001_extracted.tsv
dx download phenotypes/summary_diagnoses/40002_extracted.tsv
dx download phenotypes/summary_diagnoses/40006_extracted.tsv
dx download phenotypes/summary_diagnoses/40013_extracted.tsv
dx download phenotypes/summary_diagnoses/41270_extracted.tsv
dx download phenotypes/summary_diagnoses/41271_extracted.tsv

dx download phenotypes/first_occurrence/first_incidence_table*.tsv

# Download ICD9 conversion mapping
dx download phenotypes/summary_diagnoses/2018_I9gem.txt

# Download the full list of individuals with WES data
dx download project-GFPBQv8J0zVvBX9XJyqyqYz1:file-GFjJ0yjJ0zVb9BK82ZQFkx0y

# Download Vital Stats:
dx download phenotypes/base_covariates/vital_stats.tsv

```

# Function for Converting ICD9 to ICD10

```{r convert ICD codes}

icd9_to_icd10 <- fread("2018_I9gem.txt",col.names = c("ICD9","ICD10","flags"))

convert_icd9_to_icd10 <- function(convert_table) {

    convert_table_merge <- merge(convert_table, icd9_to_icd10, by = "ICD9", all.x = T)
    
    failed_convert_table_merge <- convert_table_merge[is.na(ICD10)]
    correct_convert_table_merge <- convert_table_merge[!is.na(ICD10)]
    
    # Now trim off the correct number of places...
    fix_failed <- function(code) {
        mapped_codes <- icd9_to_icd10[str_starts(ICD9, code)]
        
        for (i in c(3:10)) {
            col_name <- paste0('str', i)
            mapped_codes[,eval(col_name):=str_trunc(string = ICD10, width = i, side = 'right', ellipsis = '')]
        }
        
        correct_code <- NULL
        for (i in c(10:3)) {
            col_name <- paste0('str', i)
            if (length(mapped_codes[,unique(get(col_name))]) == 1) {
                correct_code <- mapped_codes[,unique(get(col_name))]
                break
            }
            
        }
        
        return(correct_code)
        
    }
    
    failed_codes <- data.table(ICD9=unique(failed_convert_table_merge[,ICD9]))
    failed_codes[,correct_code:=fix_failed(ICD9), by = 1:nrow(failed_codes)]
    
    # And merge back in...
    failed_convert_table_merge <- merge(failed_convert_table_merge, failed_codes, by = "ICD9")[,c('ICD9', 'eid','correct_code', 'flags')]
    setnames(failed_convert_table_merge, 'correct_code', 'ICD10') # I do it this way because ICD10 is still in the failed_cc table
    
    # Get unique codes
    correct_convert_table_merge[,key:=paste(ICD9, eid, sep = "_")]
    correct_convert_table_merge[,dummy:=1]
    counts <- correct_convert_table_merge[,sum(dummy),by="key"]
    
    # Spit out code/individual pairs that mapped to more than one ICD10 code
    duplicate_convert_table_merge <- unique(correct_convert_table_merge[key %in% counts[V1 > 1,key], c("ICD9","eid")])
    correct_convert_table_merge <- correct_convert_table_merge[key %in% counts[V1 == 1,key]]
    
    # Now try and rescue some of the codes by using the same 'walking up the ICD10 tree' logic that we use above for failed codes
    failed_codes <- data.table(ICD9=unique(duplicate_convert_table_merge[,ICD9]))
    failed_codes[,ICD10:=fix_failed(ICD9), by = 1:nrow(failed_codes)]
    
    # Merge bac in
    duplicate_convert_table_merge <- merge(duplicate_convert_table_merge, failed_codes, by = "ICD9")[,c('ICD9', 'eid','ICD10')]
    
    # Now put everything back together...
    translated_convert_table <- rbind(correct_convert_table_merge[,c('ICD9','eid','ICD10')], # Codes that merged with no problems
                                failed_convert_table_merge[,c('ICD9','eid','ICD10')], # Codes that didn't map at all
                                duplicate_convert_table_merge[,c('ICD9','eid','ICD10')]) # Codes that found multiple ICD10 codes
    
    # Give summary number:
    n_failed <- nrow(translated_convert_table[is.na(ICD10)])
    n_total <- nrow(convert_table)
    cat(paste0('Prop. of failed codes after rescue protocol: ', 
               sprintf(fmt = '%0.2f%%', ((n_failed / n_total) * 100)),
               ' (', n_failed, ' / ', n_total, ')'))
    
    return(translated_convert_table)
}

```

# Import Files

## First Occurrence

### Import FO Data

```{r import FO}

# Have to import as character
file_nums = c(1:12)

first_occurrence <- data.table()

for (num in file_nums) {
    
    curr_table <- fread(paste0('first_incidence_table',num,'.tsv'),colClasses = 'character')
    curr_table <- data.table(pivot_longer(curr_table,cols = names(curr_table)[2:ncol(curr_table)],names_to='code',values_drop_na=TRUE))
    first_occurrence <- rbind(first_occurrence, curr_table)
    
}

setnames(first_occurrence, 'code', 'ICD10')

```

### Add Age at Incidence

```{r}

max_date <- vital_stats[,max(dod,na.rm = T)]

vital_stats <- fread("vital_stats.tsv", colClasses = c('eid' = 'character'))
vital_stats[,current_age:=ifelse(is.na(dod),
                                 time_length(difftime(max_date, dob), "years"),
                                 time_length(difftime(dod, dob), "years"))]

# Merge first_occurance data special codes and replace with DoB for most cases:
first_occurrence <- merge(first_occurrence, vital_stats, by = 'eid', all.x=T)

# Correct DoB stats
first_occurrence <- first_occurrence[value != 'Code has event date in the future and is presumed to be a place-holder or other system default']
first_occurrence[,value:=ifelse(value == "Code has event date after participant's date of birth and falls in the same calendar year as date of birth",dob,
                                ifelse(value == "Code has event date matching participant's date of birth", dob, value))]
first_occurrence[,value:=as.Date(value)] # Have to finally convert to date now that we have fixed these issues

# Drop codes after the last death registry refresh (not many):
first_occurrence <- first_occurrence[value <= max_date]

# Drop codes that occur after patient death (not many):
first_occurrence <- first_occurrence[is.na(dod) | value <= dod]

# Calculate age at incidence:
first_occurrence[,age_at_incidence:=time_length(difftime(value, dob), "years")]
```

## HES

### Convert to Sparse Format

You will need to install the 'reticulate' R package to enable this code block to actually function. Oh, and also pray to whatever god you believe in.

```{python process HES, include=FALSE}

import ast

with open('41270_extracted.tsv', 'r') as hes:
    icd_output = open('41270_processed.tsv', 'w')
    for line in hes:
        line = line.rstrip()
        data = line.split('\t')
        eid = data[0]
        if data[1] != 'p41270' and data[1] != 'NA':
            codes = ast.literal_eval(data[1])
            for code in codes:
                written = icd_output.write(f'{eid}\t{code}\n')
            
icd_output.close()

with open('41271_extracted.tsv', 'r') as hes:
    icd_output = open('41271_processed.tsv', 'w')
    for line in hes:
        line = line.rstrip()
        data = line.split('\t')
        eid = data[0]
        if data[1] != 'p41271' and data[1] != 'NA':
            codes = ast.literal_eval(data[1])
            for code in codes:
                written = icd_output.write(f'{eid}\t{code}\n')
            
icd_output.close()

```

### Import

```{r import HES}

hes_ICD10 <- fread("41270_processed.tsv", col.names = c("eid","ICD10"), colClasses = c("character","character"))
hes_ICD9 <- fread("41271_processed.tsv", col.names = c("eid","ICD9"), colClasses = c("character","character"))

# Convert ICD9 -> 10
converted_hes_ICD9 <- convert_icd9_to_icd10(hes_ICD9)

hes <- unique(rbind(hes_ICD10, converted_hes_ICD9[!is.na(ICD10),c("eid","ICD10")]))
```

## Death Records

```{r import death records}

primary_dr <- fread("40001_extracted.tsv", colClasses = 'character')
primary_dr <- data.table(pivot_longer(primary_dr, cols = -eid,values_drop_na = T))[,c("eid","value")]
setnames(primary_dr,'value','ICD10')

secondary_dr <- fread("40002_extracted.tsv", colClasses = 'character')
secondary_dr <- data.table(pivot_longer(secondary_dr, cols = -eid,values_drop_na = T))[,c("eid","value")]
setnames(secondary_dr,'value','ICD10')

dr <- rbind(primary_dr, secondary_dr)
```

## Cancer Codes

```{r import cancer codes}

cc_ICD10 <- fread("40006_extracted.tsv", colClasses = 'character')
cc_ICD10 <- unique(data.table(pivot_longer(cc_ICD10,cols = -eid, values_drop_na = T))[,c("eid","value")])
setnames(cc_ICD10,'value','ICD10')

cc_ICD9 <- fread("40013_extracted.tsv", colClasses = 'character')
cc_ICD9 <- unique(data.table(pivot_longer(cc_ICD9,cols = -eid, values_drop_na = T))[,c("eid","value")])
setnames(cc_ICD9,'value','ICD9')

# Convert ICD9 -> 10
converted_cc_ICD9 <- convert_icd9_to_icd10(cc_ICD9)

# Mash together ICD9/ICD10
cc <- unique(rbind(cc_ICD10, converted_cc_ICD9[!is.na(ICD10),c("eid","ICD10")]))

```

# Make Database

Tables to merge:

* First Occurrence (`first_occurrence`) - considered PRIMARY record and is the basis for all other records
* Hospital Episode Stats (`hes`) - Merged ICD10 & ICD9 records
* Death Records (`dr`) - Merged Primary and Secondary causes of death (no ICD9 records)
* Cancer Codes (`cc`) - Merged ICD10 & ICD9 records

## Ingest Samples

```{r}

# Read in all WES samples
individuals <- fread('wes_samples.470k.txt', colClasses = 'character', col.names = 'eid')

# Calculate number of records per-person (unsure if actually important...)
add_N <- function(table_to_use, name) {
    N_counts <- table_to_use[,.N,by=eid]
    merged <- merge(individuals, N_counts, by = 'eid', all.x = T)
    merged[,N:=ifelse(is.na(N), 0, N)]
    setnames(merged, 'N', name)
    
    return(merged)
}

individuals <- add_N(first_occurrence, 'N_fo')
individuals <- add_N(hes, 'N_hes')
individuals <- add_N(dr, 'N_dr')
individuals <- add_N(cc, 'N_cc')

individuals[,N_recs:=sum(N_fo,N_hes,N_dr,N_cc), by = 1:nrow(individuals)]

```

## Merge FO Data

Logic:

- Go through EVERY FO record and ask if it exists in any of the other three tables.
- If the record exists, check if it is lower on the tree (i.e. longer code) and replace it in the FO data

```{r}

search_table <- unique(rbind(first_occurrence[,c("eid","ICD10")], 
                             hes[,c("eid","ICD10")], 
                             dr[,c("eid","ICD10")], 
                             cc[,c("eid","ICD10")]))

fwrite(search_table,"search_table.tsv.gz",sep="\t",col.names=T, row.names = F, quote = F, na="NA")


```









